Implementing simplified INLA into `aghq`

### 2022-10-17 src/dev_sinla/sinla.Rmd
Improve exposition and fix LHS/RHS discrepancy (fingers crossed)

```{r}
cbpalette <- multi.utils::cbpalette()
```

@@

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by exploration of $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
where $H_k(z_d)$ are the Hermite polynomials for $k \in \mathbb{N}$, and
$$
w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ with nodes $z$, of which there will be $k^d$.
Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

@@

Now, we want to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
$$
\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
$$
The goal of doing this is that the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
This would be an accurate approach, but not computationally feasible for large models.
What we hope to achieve is an approximation which gets to most of the accuracy benefits of this new Laplace approximation approach, but without the high computational cost.

@@

We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim).
`sim_data` contains many replicates of simulated data, so here we just take the first one, as given by:

@@

@@

compile("model1.cpp")
dyn.load(dynlib("model1"))
```

We create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`):

```{r}
h <- MakeADFun(
  data = dat,
  parameters = param,
  random = "phi_prev",
  DLL = "model1",
  silent = TRUE
)
```

Fit the model using MCMC:

@@

What does the posterior marginal $p(\phi_1 \, | \, y)$ look like?

```{r}
plot <- extract(mcmc, pars = "phi_prev[1]") %>%
  as.data.frame() %>%
  ggplot(aes(x = phi_prev.1.)) +
    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior")

@@

@@

The `aghq` package currently approximates latent field posterior marginals by taking marginals of the joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ (this is done implicitly, via sampling).
This is same approach used by `method = "gaussian"` in `R-INLA`.

The Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
$$
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \mu(\theta)} = \frac{p(y, \mu(\theta), \theta)}{\det \left(\frac{1}{2 \pi} Q(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
\end{equation}
$$
such that one easy way to calculate it is to use set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.
Let's optimise `h`:

@@

The values `opt$par` below are the values of $\theta$ which maximise $\tilde p(\theta \, | \, y)$:

@@

We can also look at the parameters (that is the mean vector and precision matrix) of the Gaussian approximation at the mode above (`opt$par`).
Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.

@@

#' Note that I'm calling this mean, but because it's a Gaussian the mean is the same as the mode
mean <- mm[h$env$random]
length(mean) #' The length is 30
mean

@@

As a check, we would like to reproduce `h` using the right hand side of Equation \@ref(eq:pthetagiveny).
To do this, we start by creating an objective function `f` (see Appendix for `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
This is what we will be using for the numerator of Equation \@ref(eq:pthetagiveny):

@@

For the `phi_prev` parameters, we compute their `mean` and `Q` using `h`.
We then we evaluate everything using `f` to get the numerator.
We multiply by -1 because `f$fn` gives the negative log-likelihood, and we want just the log-likelihood.
In the denominator, we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.
On the log-likelihood scale, the number we get is:

@@

theta <- param
theta$phi_prev <- NULL

@@

As compared with what we got from `h`.
They're not the same!
Something is wrong here:

```{r}
-1 * h$fn(input_hyper)
```

>>>>

```{r echo=FALSE}
cbpalette <- multi.utils::cbpalette()
```

@@

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
$$
w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
be the associated weights.
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
Then $\theta(z) = \hat \theta + Lz$ where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

@@

The goal of this notebook is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
$$
\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
$$
The reasons for doing this is that the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
This would be an accurate approach, but not computationally feasible for large models.
What we hope to achieve is an approximation which gets to most of the accuracy benefits of the Laplace approximation approach, but without the high computational cost.

@@

We begin by loading in testing data, as generated by [`prev-anc-art_sim`](https://github.com/athowes/elgm-inf/tree/master/src/prev-anc-art_sim).
For more information about this model, see the [Prevalence, ANC, ART model inference case-study](https://athowes.github.io/elgm-inf/prev-anc-art.html).
`sim_data` contains many replicates of simulated data.
Here we only need one, so will will just take the first as given by:

@@

```{r}
compile("model1.cpp")
dyn.load(dynlib("model1"))
```

We create an objective function `h` (see Appendix for `TMB` template) integrating the random effects $x$ (which here correspond to the spatial random effects `phi_prev`).
The values of `param` intitialise the objective function:

@@

h <- MakeADFun(
  data = dat,
  parameters = param,
  random = "phi_prev",
  DLL = "model1",
  silent = TRUE
)
```

It is simple to fit the model using MCMC via the `tmbstan` package.
We use 4 `chains` each of 5000 `iter`:

@@

The posterior marginal $p(\phi_1 \, | \, y)$ can be examined by extracting the corresponding Markov chains, and plotting a histogram of the draws:

```{r message=FALSE, warning=FALSE}
plot <- extract(mcmc, pars = "phi_prev[1]") %>%
  as.data.frame() %>%
  ggplot(aes(x = phi_prev.1.)) +
    geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
    theme_minimal() +
    labs(x = "phi_prev[1]", y = "Posterior")

@@

Now we will compute Equation \@ref(eq:eq:pxgiveny) taking the marignals of a joint Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$.
This is the approach currently used in the `aghq` package (done implicitly, via sampling).
This approach is also used by `method = "gaussian"` in `R-INLA`.

@@

The Gaussian approximation $p_\text{G}(x \, | \, \theta, y)$ is previously used in computing the Laplace approximation to $p(\theta \, | \, y)$
$$
\begin{equation}
\tilde p(\theta \, | \, y) \propto \frac{p(x, \theta, y)}{p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \mu(\theta)} = \frac{p(y, \mu(\theta), \theta)}{\det \left(\frac{1}{2 \pi} Q(\theta) \right)^{1/2}}, (\#eq:pthetagiveny)
\end{equation}
$$
such that one easy way to calculate it is to set `random = "x"` in the `TMB` template to integrate the latent field out, then recover the Gaussian approximation from there, as we have done in the function `h` above.

Let's optimise `h` to obtain estimates $\hat \theta$ which maximise $\tilde p(\theta \, | \, y)$:

@@

The values `opt$par` below correspond to $\hat \theta$:

@@

We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt$par` above.
Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.

@@

#' Note that I'm calling this mean, but because it's a Gaussian the mean is the same as the mode
mean <- mm[h$env$random]
length(mean) #' The length is 30

@@

As a check, we would like to reproduce `h` using the right hand side of Equation \@ref(eq:pthetagiveny).
To do this, we start by creating an objective function `f` (see Appendix for the `TMB` template) to evaluate $- \log p(x, \theta, y)$ (without any Laplace approximations!).
This is what we will be using for the numerator of Equation \@ref(eq:pthetagiveny):

@@

Given $\theta$, we compute the `mean` and `Q` of $x$ (corresponding to `phi_prev`) using `h`.
We then concatenate and evaluate using `f` to get the numerator, multiplying by -1 because `f$fn` gives the negative log-likelihood, and we want the log-likelihood.
In the denominator, we use `determinant` on the log-scale as it is more numerically stable than e.g. `log(det(...))`.

@@

theta <- param
theta$phi_prev <- NULL #' Setting the value of x to be NULL

@@

As compared with what we got from `h`:

```{r}
-1 * h$fn(unlist(theta))
```
